---
title: "Class 018 - Mini Project"
author: "Ricardo Carias (PID: A18573289)"
format: pdf
toc: TRUE
---

## Background 
Pertussis (Whooping Cough) is a common lung infection caused by the bacteria B. pertussis.
Most incompetent individuals will clear the infection. However, this infection can be deadly in infants. 

As the Center for Disease Control (CDC) tracks the number of reported cases int he US: 

Link: https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html

We can 'scrape' this data with the **datapasta** package. 

```{r}
cdc <- data.frame(
  year = c(
    1922L,
    1923L,
    1924L,
    1925L,
    1926L,
    1927L,
    1928L,
    1929L,
    1930L,
    1931L,
    1932L,
    1933L,
    1934L,
    1935L,
    1936L,
    1937L,
    1938L,
    1939L,
    1940L,
    1941L,
    1942L,
    1943L,
    1944L,
    1945L,
    1946L,
    1947L,
    1948L,
    1949L,
    1950L,
    1951L,
    1952L,
    1953L,
    1954L,
    1955L,
    1956L,
    1957L,
    1958L,
    1959L,
    1960L,
    1961L,
    1962L,
    1963L,
    1964L,
    1965L,
    1966L,
    1967L,
    1968L,
    1969L,
    1970L,
    1971L,
    1972L,
    1973L,
    1974L,
    1975L,
    1976L,
    1977L,
    1978L,
    1979L,
    1980L,
    1981L,
    1982L,
    1983L,
    1984L,
    1985L,
    1986L,
    1987L,
    1988L,
    1989L,
    1990L,
    1991L,
    1992L,
    1993L,
    1994L,
    1995L,
    1996L,
    1997L,
    1998L,
    1999L,
    2000L,
    2001L,
    2002L,
    2003L,
    2004L,
    2005L,
    2006L,
    2007L,
    2008L,
    2009L,
    2010L,
    2011L,
    2012L,
    2013L,
    2014L,
    2015L,
    2016L,
    2017L,
    2018L,
    2019L,
    2020L,
    2021L,
    2022L,
    2023L,
    2024L,
    2025L
  ),
  cases = c(
    107473,
    164191,
    165418,
    152003,
    202210,
    181411,
    161799,
    197371,
    166914,
    172559,
    215343,
    179135,
    265269,
    180518,
    147237,
    214652,
    227319,
    103188,
    183866,
    222202,
    191383,
    191890,
    109873,
    133792,
    109860,
    156517,
    74715,
    69479,
    120718,
    68687,
    45030,
    37129,
    60886,
    62786,
    31732,
    28295,
    32148,
    40005,
    14809,
    11468,
    17749,
    17135,
    13005,
    6799,
    7717,
    9718,
    4810,
    3285,
    4249,
    3036,
    3287,
    1759,
    2402,
    1738,
    1010,
    2177,
    2063,
    1623,
    1730,
    1248,
    1895,
    2463,
    2276,
    3589,
    4195,
    2823,
    3450,
    4157,
    4570,
    2719,
    4083,
    6586,
    4617,
    5137,
    7796,
    6564,
    7405,
    7298,
    7867,
    7580,
    9771,
    11647,
    25827,
    25616,
    15632,
    10454,
    13278,
    16858,
    27550,
    18719,
    48277,
    28639,
    32971,
    20762,
    17972,
    18975,
    15609,
    18617,
    6124,
    2116,
    3044,
    7063,
    22538,
    21996
    
  )
)

```

>Q. can we make a plot of these data?

```{r}


library(ggplot2)

ggplot(cdc, aes(x = year, y = cases)) +
  geom_line() +
  geom_vline(xintercept = 1946, lty = 2, col = "blue") +
  geom_vline(xintercept = 1996, lty = 2, col = "red") +
  geom_vline(xintercept = 2020, lty = 2, col = "green")

  
 
```

>Q. Add some major milestones including the first wP vaccine (1946), the aP vaccine(1996), and COVID pandemic start at (2020).  


Blue: wP vaccine introduced.

Red: Public opinion of vaccine declines and we see susceptible populations infection rates rise. Notably, around the 1996 time point, we have the rise of public access to internet and spread of misinformation.

Green: Covid_19 Pandemic 

aP induced immunity weans faster than wP immunity. It appears that immunizations which stimulate a more robust immune response, due to the full set of inactivated components present, result in stronger immunological memory.

##CMI-PB Project (Computational Models of Immunity - Pertussis Boost)

**This project aims to provide the scientific community with this very information.**

Link: https://www.cmi-pb.org/ 

They make their data available via JSON format returning API. We an read this in R with the `read_jason()` from the **jsonlite** package:
```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)

head(subject)
```

>Q. How many wP and aP individuals are in this `subject` table?

```{r}
table(subject$infancy_vac)
```

>Q. Break down of biological sex?

```{r}
table(subject$biological_sex)
```

>Q. In terms of race and biological sex, is this dataset representative of US population?

```{r}
table(subject$race, subject$biological_sex)

```



Let's read some more database tables: 

```{r}
library(jsonlite)

specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)

ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)

head(specimen)
```

```{r}
head(ab_titer)

```

To analyze this data we first need to merge these data frames / tables.

We can the `*_join()` family of functions from **dplyr** to do this. 


```{r}
library(dplyr)

meta <- inner_join(subject, specimen)

head(meta)
```

```{r}
abdata <- inner_join(meta, ab_titer)
head(abdata)
```

>Q. What antibody isotypes are measured?

```{r}
table(abdata$isotype)
```


>Q. What antigens are reported?

```{r}
table(abdata$antigen)

```

Focusing on IgG isotype and make a plot of MFI_normalized for all antigens. 





```{r}
library(dplyr)

igg <- abdata|>
  filter(isotype == "IgG")

head(igg)

```


```{r}
ggplot(igg) + 
  aes(MFI_normalised, antigen) + 
  geom_boxplot()
```


>Q. Is there a difference for aP vs wP with these values?


```{r}

ggplot(igg) + 
  aes(MFI_normalised, antigen) + 
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

```{r}
ggplot(igg) + 
  aes(MFI_normalised, antigen, col = infancy_vac) + 
  geom_boxplot() 
```

>Q. Is there a temporal response?

```{r}

ggplot(igg) + 
  aes(MFI_normalised, antigen, col = infancy_vac) + 
  geom_boxplot() +
  facet_wrap(~visit)
```


## Focus on Pretussis Toxin (PT)

```{r}
pt.igg.21 <- igg |> filter (antigen =="PT",
               dataset== ("2021_dataset")
               )
head(pt.igg.21)
```


```{r}
ggplot(pt.igg.21) +
  aes(x=planned_day_relative_to_boost,y= MFI_normalised, col=infancy_vac,
      group = subject_id) +
  geom_point() +
  geom_line() + 
  geom_vline(xintercept = 14, lty=2 , col= "black")

```


In the lab sheet there is one plot to try at home : 

its actually in the slides actually. 


```{r}
ggplot(pt.igg.21) +
  aes(x = planned_day_relative_to_boost,
      y = MFI_normalised,
      col = infancy_vac,
      group = subject_id) +
  geom_point() +
  geom_line(alpha = 0.3) +
  geom_vline(xintercept = 14, lty = 2, col = "black") +
  geom_vline(xintercept = 0, lty = 2, col = "black") +
  geom_smooth(aes(group = infancy_vac), se = FALSE)
```

