---
title: "Class 012"
author: "Ricardo Carias PID: A18573289"
format: pdf
toc: TRUE
---

## Section 1


> Q5: What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are homozygous for the asthma associated SNP (G|G)?


```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)

```


```{r}
table(mxl$Genotype..forward.strand.)
```

Nine are homozygous for the asthma associated SNP.

Towards Proportion: 

```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl) *100
```
**Therefore**, `14.0625%` are homozygous for the asthma associated SNP.


```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)

table(gbr$Genotype..forward.strand.)/nrow(gbr) *100
```

**Therefore**, `29.67033%` are homozygous for the asthma associated SNP within the GBR population.

Interpretation, the variant that is associated with childhood asthma is more frequent in GBR pop compared to MXL pop.

Let's keep digging. 

Note: The course videos, lab sheets, instructions, and other supporting documents do not align directly. You get what you get and I dont know why there is two assignements for this. 1 for lab and 1 for hw but they both have question 11 and 12.  

## Section 4 Population Analysis 

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt", header=TRUE)

head(expr)

nrow(expr)
```

 >Q13: Sample size per genotype

```{r}
table(expr$geno)
```



 >Q13: Median expression per genotype

```{r}
tapply(expr$exp, expr$geno, median)
```
There are 462 individuals which we have the data for. Distribution of genotypes: 108 A/A, 233 A/G, and 121 G/G. 

 >Q14: Boxplot

```{r}
library(ggplot2)
ggplot(expr) + 
  aes(geno, exp, fill = geno) + 
  geom_boxplot(notch = TRUE)
```
When compared across the three genotypes, we can see that the A/A genotype results in the highest median expression levels. The heterozygous A/G genotype shows intermediate expression levels, while the G/G genotype results in the lowest expression levels. This dosage affect could be explained by which alleles a given individual has rather than simple dominance. 


